The hydra and the rooster

### 2020-07-03 latex/chapter-hydras.tex
redaction Finite_ordinals

Let $B$ be some instance of class \texttt{Battle}. It is easy to define inductively the relation between the $i$-th and the $j$-th steps of a battle of type $B$.

@@

@@

\section{Reasoning about any battle}

\section{Termination}

In this section, we are interested in proofs of termination of all battles of a  given kind, and specifically proofs by \emph{variants}.

>>>>

Let $B$ be any instance of class \texttt{Battle}. It is easy to define inductively the relation between the $i$-th and the $j$-th steps of a battle of type $B$.

@@

\section{Generic properties}

Let $B$ be an instance of \texttt{Battle}. We say that $B$ is \emph{alive} if
for any configuration $(i,h)$, where $h$ is not a head, there exists a further step in class $B$.

\subsection{Termination}

@@

### 2020-07-03 latex/chapter-skippable.tex
Finite_ordinals added
redaction Finite_ordinals

\chapter{Introduction to ordinal numbers and the Coq proof assistant}

In this chapter, we introduce on a simple example the main concepts which are developped later. We assume the reader to have a very basic experience with the \coq{} proof assistant, for instance with one of the numerous tutorial one can consult on Internet
(see for instance \url{https://coq.inria.fr/documentation}).

@@

Beginners in \coq{} often play with natural numbers (the type \texttt{nat} of \coq's 
standard library), so we considered the ordinal  $\omega$ as ``well-known'' and chosed 
$\omega^2$ as a basis for our examples.

@@

\item The segment of finite ordinals is isomorphic to the set of natural numbers.
 The first infinite ordinal is the limit ordinal $\omega$.

@@

\subsection{Ordinal notations}
In a proof assistant like \coq{},  it is useful to represent ordinals through some data-type, and make arithmetical operations and comparison effectively implemented  through certified functions.

Of course, since the set of all countable ordinals is not countable, it cannot be represented by a language of finite terms.

\subsubsection{Notations systems for small ordinals}

The following list shows five popular notation systems for several ordinals (shown in increasing order).
Please keep in mind that a notation system for a given ordinal $\alpha$ represents 
ordinals \emph{strictly below} $\alpha$.

>>>>

\chapter{Introduction to ordinal numbers with the Coq proof assistant}

In this chapter, we introduce on  simple examples the main concepts which are developped later.  We assume the reader to have a very basic experience with the \coq{} proof assistant, for instance with one of the numerous tutorials one can consult on Internet
(see for instance \url{https://coq.inria.fr/documentation}). A basic experience  with Peano numbers (library \texttt{Arith}) would be useful.

@@

@@

\item Let $\alpha$ be any ordinal, the \emph{segment} $\mathbb{O}_\alpha$ is the
interval $[0,\alpha[\,=\,\{\beta|0\leq\beta<\alpha\}$.

\item The set  of finite ordinals is a segment isomorphic to the set of natural numbers.
 The first infinite ordinal is the limit ordinal $\omega$.

@@

Without ambiguity we will denote the segment $\mathbb{O}_\alpha$ by just $\alpha$.

\section{Ordinal notations}
In a proof assistant like \coq{},  it is useful to represent ordinals through some data-type, and make arithmetical operations and comparison effectively implemented  through certified functions.

Of course, since the set of all countable ordinals is not countable, it cannot be represented by a language of finite terms. So, we will study ordinal notations just  for countable segments.

\subsection{Notation systems for some small ordinals}

The following enumeration  shows  popular notation systems for several ordinals (shown in increasing order with respect to inclusion), and their possible representation in \coq{}.
Please keep in mind that a notation system for a given ordinal $\alpha$ represents 
ordinals \emph{strictly below} $\alpha$.

\subsubsection{Finite ordinals}
Let $n$ be some natural number. The segment associated with $n$ is the interval 
$[0,n[\,=\,\{0,1,\dots,n-1\}$.

One may represent the ordinal $n$ by a sigma type.

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.Prelude.Finite_ordinals.html}{Prelude.Finite\_ordinals}}

\begin{Coqsrc}
Coercion is_true: bool >-> Sortclass.

Definition t (n:nat) := {i:nat | Nat.ltb i  n}.
\end{Coqsrc}

The order on type \texttt{t $n$} is defined through the projection on \texttt{nat}.

\begin{Coqsrc}
Definition lt {n:nat} : relation (t n) :=
  fun alpha beta => Nat.ltb ( proj1_sig alpha) (proj1_sig beta).
\end{Coqsrc}

For instance, let us build two elements of the segment $[0, 7[$, \emph{i.e.} two
inhabitants of   type (\texttt{t 7}), and prove a simple  inequality (see Fig.~\ref{fig:O7}).

\begin{figure}[h]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (N0) at (0,0) {$\bullet$};
\node (i0) at (0,1) {$0$};
\node (N1) at (2,0) {$\bullet$};
\node (i1) at (2,1) {$1$};
\node (N2) at (4,0) {$\bullet$};
\node (i2) at (4,1) {$2$};
\node (N3) at (6,0) {$\bullet$};
\node (i3) at (6,1) {$3$};
\node (N4) at (8,0) {$\bullet$};
\node (i4) at (8,1) {$4$};
\node (N5) at (10,0) {$\bullet$};
\node (i5) at (10,1) {$5$};
\node (N6) at (12,0) {$\bullet$};
\node (i6) at (12,1) {$6$};
\node(alpha1) at (4,-1) {$\alpha_1$};
\node(alpha2) at (10,-1) {$\beta_1$};
\end{tikzpicture}

\caption{The segment $\mathbb{O}_7$\label{fig:O7}}
\end{figure}

\index{Coq!Commands!Program Definition}

\begin{Coqsrc}
Program Example alpha1 : t 7 := 2.

Program Example beta1 : t 7 := 5.

Example i1 : lt  alpha1 beta1.
Proof.   now compute. Qed.
\end{Coqsrc}

Note that the type \texttt{t 0} is empty, and that \texttt{Program} generates an obligation
for verifying that the constraint $i<n$ is fullfilled when one tries to build the $i$-th ordinal in type \texttt{t $n$}.

\begin{Coqsrc}
Lemma t0_empty (alpha: t 0): False.
Proof.
  destruct alpha.
  destruct x; cbn in i; discriminate.
Qed.

Program Definition bad : t 10 := 10.
Next Obligation.
  compute.
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 162)

  ============================
  false = true
\end{Coqanswer}

\begin{Coqsrc}
Abort.
\end{Coqsrc}

Note also that attempting to compare an ordinal of type \texttt{t $n$}  with an ordinal of
type \texttt{t $p$}  leads to an error if $n$ and $p$ are not convertible.

\begin{Coqsrc}

Program Example gamma1 : t 8 := 7.

Fail Goal lt alpha1 gamma1.
\end{Coqsrc}

\begin{Coqanswer}
 The command has indeed failed with message:
The term "gamma1" has type "t 8" while it is expected to have type "t 7".
\end{Coqanswer}
